The role of viscosity on drop impact forces

### 2024-08-28 Main/ViscousDropImpact_v7.tex
feat: Explain the second impact force peak in the drop impact dynamics
Provide a detailed explanation for the formation mechanism of the second impact
force peak observed in drop impact experiments and numerical simulations. The
key points are:

- The second peak is also driven by momentum conservation as the drop takes off
  the surface, coinciding with the formation of a Worthington jet.
- The time at which the second peak occurs scales with the inertio-capillary
  timescale and is insensitive to the Weber and Ohnesorge numbers.
- For low Ohnesorge and high Weber numbers, the drop retraction

We leverage the axial symmetry of the drop impact (figure~\ref{fig:schematic}a). This axial symmetry breaks at large $We$ ($\ge 100$ for water drops and even larger Weber number for more viscous drops), owing to destabilization by the surrounding gas after splashing \citep{xu2005drop, Eggers2010, Driscoll2011, riboux2014experiments, Josserand2016, zhang2022impact}. 
To solve the governing equations~\eqref{eqn:continuity}--\eqref{eqn:VoF}, the velocity field $\boldsymbol{v}$ and time $t$ are normalized by the inertio-capillary scales, $V_\gamma = \sqrt{\gamma/\rho_dD_0}$ and $\tau_\gamma = \sqrt{\rho_dD_0^3/\gamma}$, respectively. Furthermore, the pressure is normalized using the capillary pressure scale $p_\gamma = \gamma/D_0$. In such a conceptualization, $Oh$ and $We$ described in \S\ref{sec:intro} uniquely determine the system. 
The Ohnesorge number based on air viscosity $Oh_a = \left(\eta_a/\eta_d\right)Oh$ and air-drop density ratio $\rho_a/\rho_d$ are fixed at $10^{-5}$ and $10^{-3}$, respectively to minimize the influence of the surrounding medium on the impact forces. 
\Rev{Lastly, we keep the Bond number ($Bo$, see~\eqref{Eq:BoDefinition}) fixed at 1 throughout the manuscript. In our system, the relevance of gravity is characterized by the dimensionless Froude number $Fr = V_0^2/gD_0 = We/Bo$ comparing inertia with gravity. Throughout this manuscript, $Fr > 1$ and gravity's role is sub-dominant compared to inertia (for detailed discussion, see appendix~\ref{app:gravity}).}
The substrate is modeled as a no-slip and non-penetrable wall, whereas vanishing stress and pressure are applied at the remaining boundaries to mimic outflow conditions for the surrounding air. The domain boundaries are far enough from the drop not to influence its impact process ($\mathcal{L}_\text{max} \gg D_0$, $\mathcal{L}_\text{max} = 8R$ in the worst case). At $t = 0$, in our simulations, we release \Rev{a spherical drop (see appendix~\ref{app:shape})} whose south pole is $0.05D_0$ away from the substrate and is falling with a velocity $V_0$. The simulations utilize adaptive mesh refinement to finely resolve the velocity, viscous dissipation, and the VoF tracer fields. A minimum grid size $\Delta = D_0/2048$ is used for this study.

@@

Similar to the mechanism leading to the formation of the first peak (\S\ref{sec:FirstPeak}), also the mechanism for the formation of this second peak is momentum conservation. As the drop takes off the surface, it applies a force on the substrate. As noted in \S\ref{sec:intro} and \citet{zhang2022impact}, this force also coincides with the formation of a Worthington jet (figure~\ref{fig:summary}iv-vi). The time $t_2$ at which the second peak is observed scales with the inertio-capillary timescale and is insensitive to $We$ and $Oh$ (figure~\ref{fig:F2Anatomy}b,c). Once again, we invoke the analogy between drop oscillation and drop impact to explain this behavior  \citep{Richard2002, chevy2012liquid}. At the time instant $t_2 \approx 0.44\tau_\gamma$, the drop's internal motion undergoes a transition from a predominantly radial flow to a vertical one due to the formation of the Worthington jet \citep{chantelot2018rebonds, zhang2022impact}. Figure~\ref{fig:F2Phenomenology} exemplifies this jet in the $Oh-We$ parameter space, which is intricately related to the second peak in the drop impact force. For low $Oh$ and large $We$, the drop retraction follows a modified Taylor-Culick dynamics \citep{bartolo2005retraction, Eggers2010, sanjay2022TC}. As $We$ is increased, the jet gets thinner but faster, maintaining a constant momentum flux $\rho_dV_j^2d_j^2$, where $V_j$ and $d_j$ are the jet's velocity and diameter, respectively \citep[figure~\ref{fig:F2Phenomenology},][]{zhang2022impact}. This invariance leads to the observed scaling $F_2 \sim F_\rho$ in this regime ($F_2 \approx 0.37F_\rho$ for $We \ge 30, Oh \le 0.01$).

@@

However, this singular jet regime is very narrow in the $Oh-We$ phase space. Figure~\ref{fig:KillSingularity}b shows two cases for water drops ($Oh = 0.0025$) at different $We$ ($5$ and $12$ for figures~\ref{fig:KillSingularity}b-i and b-ii, respectively). Bubble entrainment does not occur in either of these cases. Consequently, the maximum force amplitude diminishes for these two cases (figure~\ref{fig:F2Anatomy}). Nonetheless, these cases are still associated with high local viscous dissipation near the axis of symmetry owing to the singular nature of the flow. Another mechanism to inhibit this singular Worthington jet is viscous dissipation in the bulk. As the Ohnesorge number increases, this singular jet formation disappears ($Oh = 0.005$, figure~\ref{fig:KillSingularity}c-i), significantly reducing the second peak of the impact force. For even higher viscosities, the drop no longer exhibits the sharp, focused jet formation seen at lower viscosities, and the second peak in the force is notably diminished ($Oh = 0.05$, figure~\ref{fig:KillSingularity}c-ii).

@@

For the first impact force peak amplitude $F_1$, owing to the momentum balance after the inertial shock at impact, figure~\ref{fig:RegimeMaps}(a) summarizes the different regimes in the $Oh--We$ phase space. For low $Oh$, inertial forces predominantly dictate the impact dynamics, such that $F_1$ scales with the inertial force $F_\rho$ \citep{Philippi2016, Gordillo2018, Mitchell2019, cheng2021drop, zhang2022impact} and is insensitive to viscosity variations up to 100-fold. As $Oh$ increases, the viscosity becomes significant, leading to a new scaling law: $F_1 \sim F_\rho\sqrt{Oh}$. The paper unravels this viscous scaling behavior by accounting for the loss of initial kinetic energy owing to viscous dissipation inside the drop. Lastly, at low $We$, the capillary pressure inside the drop leads to the scaling $F_1 \sim F_\gamma$ \citep{molavcek2012quasi, chevy2012liquid}.

The normal reaction force described in this work is responsible for deforming the drop as it spreads onto the substrate, where it stops thanks to surface tension. If the substrate is non-wetting, it retracts to minimize the surface energy and finally takes off \citep{Richard2000}. In this case, the momentum conservation leads to the formation of a Worthington jet and a second peak in the normal reaction force, as summarized in figure~\ref{fig:RegimeMaps}(b). For low $Oh$ and high $We$, the second force peak amplitude scales with the inertial force ($F_\rho$), following a modified Taylor-Culick dynamics \citep{Eggers2010}. In contrast, capillary forces dominate at low $We$ and low $Oh$, leading to a force amplitude scaling of $F_2 \sim F_\gamma$. We also identify a narrow regime in the $Oh--We$ phase space where a singular Worthington jet forms, significantly increasing $F_2$ \citep{Bartolo2006Singular, zhang2022impact}, localized in the parameter space for $We \approx 9$ and $Oh < 0.01$. As $Oh$ increases, bulk viscous dissipation counteracts this jet formation, diminishing the second peak and ultimately inhibiting drop bouncing.

Our findings have far-reaching implications, not only enriching the fundamental understanding of fluid dynamics of drop impact but also informing practical applications in diverse fields such as inkjet printing, public health, agriculture, and material science where the entire range of $Oh-We$ phase space is relevant (figures~\ref{fig:schematic}b and~\ref{fig:RegimeMaps}). While this has identified new scaling laws, it also opens avenues for future research. For instance, it would be interesting to use the energy accounting approach to unify the scaling laws for the maximum spreading diameter for arbitrary $Oh$ \citep{laan2014maximum, Wildeman2016}. Although, the implicit theoretical model summarized in \citet{cheng2021drop} describes most of data in figure~\ref{fig:F1Anatomy_2}, we stress the importance of having a predictive model to determine $F_1$ for given $We$ and $Oh$ \citep{sanjay2024PRL}. The $We$ influence on the impact force also warrants further exploration, especially in the regime $We \ll 1$ for arbitrary $Oh$ \citep{chevy2012liquid, molavcek2012quasi} and drop impact on compliant surfaces \citep{alventosa_cimpeanu_harris_2023, ma2023scaling}. Another potential extension of this work is to non-Newtonian fluids \citep{martouzet2021dynamic, aguero2022impact, bertin2023similarity, jin2023marbles}.\\[5mm]

>>>>

We leverage the axial symmetry of the drop impact (figure~\ref{fig:schematic}a). This axial symmetry breaks at large $We$ ($\ge 100$ for water drops and even larger Weber number for more viscous drops), owing to destabilization by the surrounding gas after splashing \citep{xu2005drop, Eggers2010, Driscoll2011, riboux2014experiments, Josserand2016, zhang2022impact}. 
To solve the governing equations~\eqref{eqn:continuity}-\eqref{eqn:VoF}, the velocity field $\boldsymbol{v}$ and time $t$ are normalized by the inertio-capillary scales, $V_\gamma = \sqrt{\gamma/\rho_dD_0}$ and $\tau_\gamma = \sqrt{\rho_dD_0^3/\gamma}$, respectively. Furthermore, the pressure is normalized using the capillary pressure scale $p_\gamma = \gamma/D_0$. In such a conceptualization, $Oh$ and $We$ described in \S\ref{sec:intro} uniquely determine the system. 
The Ohnesorge number based on air viscosity $Oh_a = \left(\eta_a/\eta_d\right)Oh$ and air-drop density ratio $\rho_a/\rho_d$ are fixed at $10^{-5}$ and $10^{-3}$, respectively to minimize the influence of the surrounding medium on the impact forces. 
\Rev{Lastly, we keep the Bond number ($Bo$, see~\eqref{Eq:BoDefinition}) fixed at 1 throughout the manuscript. In our system, the relevance of gravity is characterized by the dimensionless Froude number $Fr = V_0^2/gD_0 = We/Bo$ comparing inertia with gravity. Throughout this manuscript, $Fr > 1$ and gravity's role is sub-dominant compared to inertia (for detailed discussion, see appendix~\ref{app:gravity}).}
The substrate is modeled as a no-slip and non-penetrable wall, whereas vanishing stress and pressure are applied at the remaining boundaries to mimic outflow conditions for the surrounding air. The domain boundaries are far enough from the drop not to influence its impact process ($\mathcal{L}_\text{max} \gg D_0$, $\mathcal{L}_\text{max} = 8R$ in the worst case). At $t = 0$, in our simulations, we release \Rev{a spherical drop (see appendix~\ref{app:shape})} whose south pole is $0.05D_0$ away from the substrate and is falling with a velocity $V_0$. The simulations utilize adaptive mesh refinement to finely resolve the velocity, viscous dissipation, and the VoF tracer fields. A minimum grid size $\Delta = D_0/2048$ is used for this study.

@@

Similar to the mechanism leading to the formation of the first peak (\S\ref{sec:FirstPeak}), also the mechanism for the formation of this second peak is momentum conservation. As the drop takes off the surface, it applies a force on the substrate. As noted in \S\ref{sec:intro} and \citet{zhang2022impact}, this force also coincides with the formation of a Worthington jet (figure~\ref{fig:summary}iv-vi). The time $t_2$ at which the second peak is observed scales with the inertio-capillary timescale and is insensitive to $We$ and $Oh$ (figure~\ref{fig:F2Anatomy}b,c). Once again, we invoke the analogy between drop oscillation and drop impact to explain this behavior  \citep{Richard2002, chevy2012liquid}. At the time instant $t_2 \approx 0.44\tau_\gamma$, the drop's internal motion undergoes a transition from a predominantly radial flow to a vertical one due to the formation of the Worthington jet \citep{chantelot2018rebonds, zhang2022impact}. Figure~\ref{fig:F2Phenomenology} exemplifies this jet in the $Oh$-$We$ parameter space, which is intricately related to the second peak in the drop impact force. For low $Oh$ and large $We$, the drop retraction follows a modified Taylor-Culick dynamics \citep{bartolo2005retraction, Eggers2010, sanjay2022TC}. As $We$ is increased, the jet gets thinner but faster, maintaining a constant momentum flux $\rho_dV_j^2d_j^2$, where $V_j$ and $d_j$ are the jet's velocity and diameter, respectively \citep[figure~\ref{fig:F2Phenomenology},][]{zhang2022impact}. This invariance leads to the observed scaling $F_2 \sim F_\rho$ in this regime ($F_2 \approx 0.37F_\rho$ for $We \ge 30, Oh \le 0.01$).

@@

However, this singular jet regime is very narrow in the $Oh$-$We$ phase space. Figure~\ref{fig:KillSingularity}b shows two cases for water drops ($Oh = 0.0025$) at different $We$ ($5$ and $12$ for figures~\ref{fig:KillSingularity}b-i and b-ii, respectively). Bubble entrainment does not occur in either of these cases. Consequently, the maximum force amplitude diminishes for these two cases (figure~\ref{fig:F2Anatomy}). Nonetheless, these cases are still associated with high local viscous dissipation near the axis of symmetry owing to the singular nature of the flow. Another mechanism to inhibit this singular Worthington jet is viscous dissipation in the bulk. As the Ohnesorge number increases, this singular jet formation disappears ($Oh = 0.005$, figure~\ref{fig:KillSingularity}c-i), significantly reducing the second peak of the impact force. For even higher viscosities, the drop no longer exhibits the sharp, focused jet formation seen at lower viscosities, and the second peak in the force is notably diminished ($Oh = 0.05$, figure~\ref{fig:KillSingularity}c-ii).

@@

For the first impact force peak amplitude $F_1$, owing to the momentum balance after the inertial shock at impact, figure~\ref{fig:RegimeMaps}(a) summarizes the different regimes in the $Oh$-$We$ phase space. For low $Oh$, inertial forces predominantly dictate the impact dynamics, such that $F_1$ scales with the inertial force $F_\rho$ \citep{Philippi2016, Gordillo2018, Mitchell2019, cheng2021drop, zhang2022impact} and is insensitive to viscosity variations up to 100-fold. As $Oh$ increases, the viscosity becomes significant, leading to a new scaling law: $F_1 \sim F_\rho\sqrt{Oh}$. The paper unravels this viscous scaling behavior by accounting for the loss of initial kinetic energy owing to viscous dissipation inside the drop. Lastly, at low $We$, the capillary pressure inside the drop leads to the scaling $F_1 \sim F_\gamma$ \citep{molavcek2012quasi, chevy2012liquid}.

The normal reaction force described in this work is responsible for deforming the drop as it spreads onto the substrate, where it stops thanks to surface tension. If the substrate is non-wetting, it retracts to minimize the surface energy and finally takes off \citep{Richard2000}. In this case, the momentum conservation leads to the formation of a Worthington jet and a second peak in the normal reaction force, as summarized in figure~\ref{fig:RegimeMaps}(b). For low $Oh$ and high $We$, the second force peak amplitude scales with the inertial force ($F_\rho$), following a modified Taylor-Culick dynamics \citep{Eggers2010}. In contrast, capillary forces dominate at low $We$ and low $Oh$, leading to a force amplitude scaling of $F_2 \sim F_\gamma$. We also identify a narrow regime in the $Oh$-$We$ phase space where a singular Worthington jet forms, significantly increasing $F_2$ \citep{Bartolo2006Singular, zhang2022impact}, localized in the parameter space for $We \approx 9$ and $Oh < 0.01$. As $Oh$ increases, bulk viscous dissipation counteracts this jet formation, diminishing the second peak and ultimately inhibiting drop bouncing.

Our findings have far-reaching implications, not only enriching the fundamental understanding of fluid dynamics of drop impact but also informing practical applications in diverse fields such as inkjet printing, public health, agriculture, and material science where the entire range of $Oh$-$We$ phase space is relevant (figures~\ref{fig:schematic}b and~\ref{fig:RegimeMaps}). While this has identified new scaling laws, it also opens avenues for future research. For instance, it would be interesting to use the energy accounting approach to unify the scaling laws for the maximum spreading diameter for arbitrary $Oh$ \citep{laan2014maximum, Wildeman2016}. Although, the implicit theoretical model summarized in \citet{cheng2021drop} describes most of data in figure~\ref{fig:F1Anatomy_2}, we stress the importance of having a predictive model to determine $F_1$ for given $We$ and $Oh$ \citep{sanjay2024PRL}. The $We$ influence on the impact force also warrants further exploration, especially in the regime $We \ll 1$ for arbitrary $Oh$ \citep{chevy2012liquid, molavcek2012quasi} and drop impact on compliant surfaces \citep{alventosa_cimpeanu_harris_2023, ma2023scaling}. Another potential extension of this work is to non-Newtonian fluids \citep{martouzet2021dynamic, aguero2022impact, bertin2023similarity, jin2023marbles}.\\[5mm]
